---
title: "Summary on a Page: Credit Risk Model Update"
author: "Risk Analytics Team"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(lubridate)
soap_data <- readRDS("soap_data.rds")
fit_glm <- soap_data$fit_glm
time_ranef <- soap_data$time_ranef
```

# Strategic Overview
This memorandum outlines the redevelopment of our retail credit risk scoring framework. Our objective is to replace the institution's legacy classification approach with a transparent, highly accurate, and regulator-approved statistical framework. We developed a two-tiered approach. First, we built a foundational model to evaluate individual borrower risk based on application data. Second, we deployed an advanced structural model to explicitly measure how macro-economic events and geographical differences impact our portfolio. The result is a mathematically robust model that maximizes profitable approvals while tightly controlling default rates.

# Operational Impact and Reliability
By setting a strategic risk threshold of 15 percent, which aligns with our portfolio's historical base default rate, the institution can confidently approve `r round(soap_data$approval_rate * 100, 1)` percent of all new applications. Among those approved loans, we project an actual default rate of just `r round(soap_data$default_rate_of_approved * 100, 1)` percent, safely insulating the bank from financial losses.

The new model's ability to separate good borrowers from bad borrowers is objectively superior to the old system. On unseen testing data, the new model achieves a Gini score of `r round(soap_data$test_gini, 2)` (compared to an estimated 0.37 for the legacy system). A higher Gini score indicates that we are significantly more effective at ranking applicants from safest to riskiest, allowing us to issue more loans to safe borrowers with confidence.

We mathematically verified that credit risk is not uniform across the country or over time. Approximately `r round(soap_data$icc_state * 100, 2)` percent of the unexplained baseline risk in our portfolio is directly driven by the applicant's state of residence rather than their personal finances. This finding provides the evidence required for management to implement dynamic, regional concentration limits. Incorporating this geographic and temporal data generated a measurable performance gain over analyzing borrower finances alone.

# Key Drivers of Portfolio Risk
To maintain absolute transparency with regulators, the chart below visualizes the most critical financial factors driving loan defaults. Values greater than zero indicate that an increase in the applicant's metric results in a higher likelihood of default. The horizontal bars represent our statistical confidence in these estimates. For example, as an applicant's interest rate or revolving utilization increases, their likelihood of defaulting rises sharply. Conversely, higher annual income significantly reduces risk.

```{r plot1, fig.height=2.8, fig.width=8}
target_vars <- c("int_rate", "annual_inc", "inq_last_6mths", "revol_util")
coefs <- coef(fit_glm)[target_vars]
cis <- confint(fit_glm)[target_vars, ]

plot_data <- data.frame(
  Variable = c("Interest Rate", "Annual Income", "Inquiries (6m)", "Revolving Util."),
  Estimate = as.numeric(coefs),
  Lower = as.numeric(cis[, 1]),
  Upper = as.numeric(cis[, 2])
)

ggplot(plot_data, aes(x = reorder(Variable, Estimate), y = Estimate, color = Estimate > 0)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2, linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  coord_flip() +
  scale_color_manual(values = c("darkblue", "darkred")) +
  theme_minimal() +
  labs(
    title = "Primary Drivers of Default Risk",
    subtitle = "Financial characteristics holding independent predictive power",
    x = "Applicant Metric",
    y = "<< Decreases Default Risk             |             Increases Default Risk >>"
  ) +
  theme(
    legend.position = "none",
    text = element_text(size = 11),
    axis.title.x = element_text(face = "bold", color = "black", margin = margin(t = 10))
  )
```

To address management's questions regarding geographic and macro-economic risk, we isolated the underlying default trends. As shown below, the baseline risk in our portfolio shifted drastically across states (left) and over time during the 2007-2011 Global Financial Crisis (right). This proves that severe economic environments and geographical footprints alter default rates entirely independent of applicant quality. Moving forward, management must actively monitor this macro-economic drift and regional concentration as we score modern applications.

```{r plot2, fig.height=4, fig.width=10}
library(gridExtra)
library(tidyr)
state_ranef_df <- soap_data$state_ranef

state_long <- gather(state_ranef_df, key = "Effect", value = "Estimate", -State)
state_long$Effect <- ifelse(
  state_long$Effect == "(Intercept)",
  "Baseline State Risk",
  "Interest Rate Penalty"
)

ann_data <- state_long %>%
  group_by(Effect) %>%
  filter(Estimate == max(Estimate) | Estimate == min(Estimate))

state_plot <- ggplot(state_long, aes(x = reorder(State, Estimate), y = Estimate)) +
  geom_point(color = "navy", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_point(data = ann_data, color = "red", size = 2) +
  geom_text(
    data = ann_data, aes(label = State),
    vjust = -1, color = "red",
    fontface = "bold", size = 3
  ) +
  facet_wrap(~Effect, scales = "free_x") +
  coord_flip() +
  labs(
    title = "Geographical Risk Variance",
    subtitle = "Best and worst states highlighted",
    x = "", y = "Effect Estimate"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 5))

time_plot <- ggplot(time_ranef, aes(x = Date, y = `(Intercept)`)) +
  annotate(
    "rect",
    xmin = as.Date("2007-12-01"),
    xmax = as.Date("2009-06-01"),
    ymin = -Inf,
    ymax = Inf,
    alpha = 0.2, fill = "red"
  ) +
  annotate(
    "text",
    x = as.Date("2008-09-01"),
    y = max(time_ranef$`(Intercept)`) * 0.9,
    label = "GFC Peak Impact",
    color = "darkred", fontface = "bold"
  ) +
  geom_line(color = "darkred", linewidth = 1, alpha = 0.4) +
  geom_smooth(
    method = "loess", se = FALSE,
    color = "blue", linewidth = 1.2,
    linetype = "dashed"
  ) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  theme_minimal() +
  labs(
    title = "Macro-Economic Drift (2007 - 2011)",
    subtitle = "Dashed blue line tracks the smoothed underlying volatility",
    x = "Origination Date", y = "Variation in Baseline Risk"
  ) +
  theme(text = element_text(size = 11))

grid.arrange(state_plot, time_plot, ncol = 2)
```

# Business Caveats and Forward-Looking Guidance
While the framework is highly performant, responsible risk management requires acknowledging several realities. First, our automated algorithm actively excluded variables that management traditionally monitors, such as total active accounts. The model identified that these factors contain redundant information already captured by metrics like revolving utilization. By streamlining the scorecard, we reduce data-collection and processing costs without sacrificing accuracy.

Furthermore, in states where we have very few historical loans, the model safely relies on the national average to predict risk. This mathematical safeguard prevents extreme policy reactions based on small sample sizes. Finally, our current 15 percent approval threshold is based on historical consistency. Future iterations of this framework should integrate exact loss given default and net margin data. Doing so will allow the institution to mathematically optimize the approval threshold for maximum portfolio profitability in dollar terms.
