---
output: 
  pdf_document:
    latex_engine: pdflatex
geometry: "top=0.38in,bottom=0.50in,left=0.55in,right=0.55in,footskip=0.20in"
fontsize: 10pt
header-includes:
  - \usepackage{booktabs}
  - \usepackage{colortbl}
  - \usepackage{xcolor}
  - \definecolor{midnightblue}{RGB}{25,25,112}
  - \definecolor{darkred}{RGB}{139,0,0}
  - \definecolor{forestgreen}{RGB}{34,139,34}
  - \usepackage{enumitem}
  - \usepackage{titlesec}
  - \usepackage{microtype}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \renewcommand{\headrulewidth}{0pt}
  - \renewcommand{\footrulewidth}{0pt}
  - \rfoot{\footnotesize\textcolor{gray}{Page \thepage\ of 2}}
  - \titlespacing*{\section}{0pt}{3pt plus 1pt minus 1pt}{1.5pt plus 1pt minus 1pt}
  - \titlespacing*{\subsection}{0pt}{2pt plus 1pt minus 1pt}{1pt plus 1pt minus 1pt}
  - \setlist[itemize]{leftmargin=*,itemsep=1.2pt,topsep=1.2pt,partopsep=0pt,parsep=0pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE, 
  fig.align = "center",
  dpi = 300
)
library(ggplot2)
library(dplyr)
library(lubridate)
library(patchwork)
library(knitr)
library(kableExtra)

soap_data <- readRDS("assets/soap_data.rds")
fit_glm <- soap_data$fit_glm
```

```{=latex}
\begin{center}
\vspace{-10pt}
{\LARGE\textbf{\textcolor{midnightblue}{SOAP: Credit Risk Model Redevelopment}}}\\[2pt]
\textbf{Prepared By: Tri Tin Vo, Connor Shannon, Omid Nodeh, Nathan Burns} $\mid$ October 2026
\end{center}
\vspace{-8pt}
\rule{\textwidth}{1pt}
\vspace{-7pt}
```

# Executive Decision and Performance Transformation
The rebuilt logistic model materially improves applicant ranking over the supplied legacy benchmark. The legacy benchmark showed weak rank discrimination (validation Gini of **0.110**, AUC of **0.555**), which was insufficiently discriminating for the proposed banking use without redevelopment and validation. On common validation data, the rebuilt Generalized Linear Model (GLM) increased Gini by **+0.286 points** to **0.396**, while held-out testing supported similar performance on comparable historical applications, with Gini of **0.374** and AUC of **0.687**.

**Management Decision:** Retain the transparent GLM as the candidate applicant-ranking model, subject to redevelopment or recalibration on contemporary Australian data and independent validation. Reserve the spatial-temporal Generalized Linear Mixed Model (GLMM) for portfolio monitoring rather than live applicant scoring.

*Methods:* Binomial logistic regression, backward AIC selection, stratified 5-fold cross-validation, held-out validation and testing, ROC/Gini and calibration assessment, and a state-month random-intercept GLMM with forward temporal validation.

```{r kpi_table}
kpi_df <- data.frame(
  `Performance Metric` = c(
    "Validation Discrimination (Gini)",
    "Held-Out Test Sample",
    "Maximum Separation (KS)",
    "Illustrative Policy Scenario (15% Cutoff)"
  ),
  `Legacy Benchmark` = c("0.110 (AUC = 0.555)", "Not available", "Not available", "Not applicable"),
  `Rebuilt Model` = c("0.396 (AUC = 0.698)", "Gini 0.374; AUC 0.687", "KS = 0.280 (at 40.1% volume)", "60.6% approval volume"),
  `Operational Takeaway` = c(
    "+0.286 validation Gini points on common evaluation sample",
    "95% CI: AUC [0.670, 0.703]; performance remained moderate on untouched historical test sample",
    "Maximum separation between cumulative default and non-default score distributions at 40.1% volume",
    "Approved default rate: 9.2% (vs 14.9% baseline); captures 62.5% of defaulting accounts"
  ),
  check.names = FALSE
)

kable(kpi_df, booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position", "striped"), font_size = 7.6) |>
  column_spec(1, bold = TRUE, width = "3.8cm") |>
  column_spec(2, width = "2.5cm") |>
  column_spec(3, width = "3.5cm") |>
  column_spec(4, width = "7.8cm")
```

# Primary Risk Drivers: Conditional Associations with Default Odds
To support explainability, the scorecard estimates conditional Odds Ratios (OR) with 95% Wald confidence intervals. Factors above 1.0 are associated with higher conditional default odds; factors below 1.0 are associated with lower conditional default odds.

```{r plot1, fig.height=1.65, fig.width=7.8, out.width="100%"}
glm_summary <- summary(fit_glm)$coefficients
ci_wald <- confint.default(fit_glm)

plot_df <- data.frame(
  Feature = c(
    "Small Business Purpose",
    "Employment: Unknown",
    "Loan Term: 60 Months",
    "Derogatory Public Record",
    "Inquiries (Last 6 Months)",
    "Interest Rate (+1.0%)",
    "Revolving Balance (+$10k)",
    "Revolving Util. (+10%)",
    "Annual Income (+$10k)"
  ),
  OR = c(
    exp(glm_summary["purposesmall_business", 1]),
    exp(glm_summary["emp_lengthUnknown", 1]),
    exp(glm_summary["term60 months", 1]),
    exp(glm_summary["pub_rec", 1]),
    exp(glm_summary["inq_last_6mths", 1]),
    exp(glm_summary["int_rate", 1]),
    exp(glm_summary["revol_bal", 1] * 10000),
    exp(glm_summary["revol_util", 1] * 0.10),
    exp(glm_summary["annual_inc", 1] * 10000)
  ),
  Lower = c(
    exp(ci_wald["purposesmall_business", 1]),
    exp(ci_wald["emp_lengthUnknown", 1]),
    exp(ci_wald["term60 months", 1]),
    exp(ci_wald["pub_rec", 1]),
    exp(ci_wald["inq_last_6mths", 1]),
    exp(ci_wald["int_rate", 1]),
    exp(ci_wald["revol_bal", 1] * 10000),
    exp(ci_wald["revol_util", 1] * 0.10),
    exp(ci_wald["annual_inc", 1] * 10000)
  ),
  Upper = c(
    exp(ci_wald["purposesmall_business", 2]),
    exp(ci_wald["emp_lengthUnknown", 2]),
    exp(ci_wald["term60 months", 2]),
    exp(ci_wald["pub_rec", 2]),
    exp(ci_wald["inq_last_6mths", 2]),
    exp(ci_wald["int_rate", 2]),
    exp(ci_wald["revol_bal", 2] * 10000),
    exp(ci_wald["revol_util", 2] * 0.10),
    exp(ci_wald["annual_inc", 2] * 10000)
  ),
  Impact = c(
    "+176% odds",
    "+73% odds",
    "+56% odds",
    "+28% odds / rec",
    "+15% odds / inq",
    "+12% odds / 1.0%",
    "+4.6% odds / $10k",
    "+4.3% odds / 10% pts",
    "-6.6% odds / $10k"
  ),
  Type = c("High Risk", "High Risk", "Risk", "Risk", "Risk", "Risk", "Risk", "Risk", "Protective")
)

plot_df$Feature <- factor(plot_df$Feature, levels = rev(plot_df$Feature))

ggplot(plot_df, aes(x = OR, y = Feature, color = Type)) +
  annotate("rect", xmin = 0.75, xmax = 1.0, ymin = -Inf, ymax = Inf, fill = "honeydew", alpha = 0.6) +
  annotate("rect", xmin = 1.0, xmax = 4.4, ymin = -Inf, ymax = Inf, fill = "mistyrose", alpha = 0.3) +
  geom_vline(xintercept = 1.0, linetype = "dashed", color = "gray40", linewidth = 0.6) +
  geom_errorbar(aes(xmin = Lower, xmax = Upper), orientation = "y", width = 0.25, linewidth = 0.75) +
  geom_point(aes(size = Type)) +
  geom_text(aes(x = Upper + 0.08, label = sprintf("OR: %.2f (%s)", OR, Impact)),
            hjust = 0, size = 2.4, fontface = "bold", show.legend = FALSE) +
  scale_color_manual(values = c("High Risk" = "firebrick3", "Risk" = "firebrick4", "Protective" = "forestgreen")) +
  scale_size_manual(values = c("High Risk" = 2.8, "Risk" = 2.0, "Protective" = 2.8)) +
  scale_x_continuous(limits = c(0.75, 4.4), breaks = c(0.8, 1.0, 1.5, 2.0, 2.5, 3.0, 3.5)) +
  labs(
    title = "Conditional Associations with Default Odds (GLM Estimates with 95% Wald Intervals)",
    subtitle = "Multiplicative scale: values > 1.0 indicate elevated default odds; values < 1.0 indicate protective buffers",
    x = "Odds Ratio (OR) Scale  [ < 1.0: Lower Estimated Odds  |  > 1.0: Default Odds Multiplier ]",
    y = NULL
  ) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 8.8, color = "midnightblue"),
    plot.subtitle = element_text(size = 6.6, color = "gray30", margin = margin(b = 2)),
    axis.text.y = element_text(face = "bold", size = 7.0, color = "black"),
    axis.title.x = element_text(face = "bold", size = 6.6, color = "gray20", margin = margin(t = 2)),
    panel.grid.minor = element_blank()
  )
```

* **Largest Positive Associations:** Small-business loans carry $+175.7\%$ higher estimated conditional default odds ($\text{OR} = 2.76$, 95% CI: $[2.13, 3.56]$) relative to car loans. Applications with unreported employment tenure have $+72.9\%$ higher odds ($\text{OR} = 1.73$, $[1.37, 2.19]$), and 60-month terms carry $+56.4\%$ higher odds ($\text{OR} = 1.56$, $[1.43, 1.71]$). *Recommendation:* Review small-business-purpose applications under a dedicated policy assessment; treat missing employment tenure as a data-quality signal before operational use.
* **Additional Risk Indicators:** Interest rate, inquiries, derogatory public records, revolving balance, and line utilization were all positively associated with default odds. Each additional derogatory public record was associated with $27.9\%$ higher estimated conditional default odds ($\text{OR} = 1.28$, $[1.12, 1.47]$).
* **Lower Estimated Odds:** Higher annual income was consistently associated with lower conditional default odds: $\text{OR} = 0.934$ per additional $\$10,000$, 95% CI: $[0.923, 0.945]$.
* **Expected-but-Dropped Account Depth Variables:** Total accounts (`total_acc`) and open accounts (`open_acc`) were not retained by backward AIC after adjustment for other candidate predictors, indicating limited incremental contribution under the selected specification. *Governance note:* Do not remove these fields operationally without local validation and assessment of their broader underwriting uses.

\newpage

# Geographic and Vintage Dynamics: The "Ranking vs. Monitoring" Paradox
State and month random effects improve portfolio description and vintage monitoring, but do not materially improve applicant ranking:

```{r plot2, fig.height=1.85, fig.width=7.8, out.width="100%"}
state_df <- soap_data$state_ranef %>% arrange(Estimate) %>% mutate(Rank = row_number())
sig_states <- state_df %>% filter(Significant == "Excludes Zero")
low_state <- state_df %>% filter(Estimate == min(Estimate))

p2_state <- ggplot(state_df, aes(x = Estimate, y = Rank, color = Significant)) +
  annotate("rect", xmin = -0.12, xmax = 0.12, ymin = -Inf, ymax = Inf, fill = "gray92", alpha = 0.5) +
  annotate("text", x = 0.0, y = 14, label = "47 of 50 Jurisdictions\nOverlap Baseline", color = "gray45", size = 2.0, fontface = "italic") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray30") +
  geom_errorbar(aes(xmin = Lower, xmax = Upper), orientation = "y", width = 0.25, linewidth = 0.35) +
  geom_point(aes(size = Significant)) +
  annotate("label", x = 0.44, y = 30,
           label = "Only 3 States Exclude Zero:\n* NV: +0.43 (Highest deviation)\n* FL: +0.32\n* CA: +0.23\n(Exploratory signals)",
           color = "firebrick4", fontface = "bold", size = 1.9, fill = "white") +
  annotate("text", x = -0.22, y = 4, label = "CO (-0.19)", fontface = "bold", size = 2.0, color = "steelblue4") +
  scale_color_manual(values = c("Excludes Zero" = "firebrick3", "Overlaps Zero" = "gray65")) +
  scale_size_manual(values = c("Excludes Zero" = 1.8, "Overlaps Zero" = 0.7)) +
  scale_x_continuous(limits = c(-0.55, 0.78), breaks = seq(-0.4, 0.6, 0.2)) +
  scale_y_continuous(breaks = NULL) +
  theme_minimal(base_size = 7.2) +
  labs(
    title = "State Baseline Deviations (GLMM)",
    subtitle = "Approximate 95% conditional intervals; state variance is 0.88% of total variance",
    y = "US States (Ranked by Risk)",
    x = "Baseline Log-Odds Deviation (0 = Model-Wide Baseline)"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 7.8, color = "midnightblue"),
    plot.subtitle = element_text(size = 6.2, color = "gray30"),
    axis.title.x = element_text(size = 6.6, face = "bold"),
    axis.title.y = element_text(size = 6.6, face = "bold"),
    plot.margin = margin(t = 2, r = 3, b = 2, l = 6, unit = "pt")
  )

time_df <- soap_data$time_ranef
p2_time <- ggplot(time_df, aes(x = Date, y = `(Intercept)`)) +
  annotate("rect", xmin = as.Date("2007-12-01"), xmax = as.Date("2009-06-01"),
           ymin = -Inf, ymax = Inf, fill = "mistyrose", alpha = 0.6) +
  annotate("rect", xmin = as.Date("2011-01-01"), xmax = as.Date("2011-12-01"),
           ymin = -Inf, ymax = Inf, fill = "#F2F6FA", alpha = 0.8) +
  annotate("text", x = as.Date("2008-09-01"), y = 0.18,
           label = "GFC Recession Period\n(NBER Dec 2007-Jun 2009)", color = "firebrick4", fontface = "bold", size = 2.0) +
  geom_vline(xintercept = as.Date("2011-01-01"), linetype = "dotted", color = "#2C5F8A", linewidth = 0.45) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_line(color = "steelblue4", linewidth = 0.5, alpha = 0.5) +
  geom_point(color = "steelblue4", size = 0.9) +
  geom_smooth(method = "loess", se = FALSE, color = "navy", linewidth = 0.8) +
  annotate("segment",
           x = as.Date("2011-07-01"), y = -0.11,
           xend = as.Date("2011-07-01"), yend = -0.01,
           arrow = arrow(length = unit(0.08, "cm")),
           linewidth = 0.35, colour = "#2C5F8A") +
  annotate("label",
           x = as.Date("2011-07-01"), y = -0.17,
           label = "2011 prospective\nholdout",
           size = 2.0, fontface = "bold",
           fill = scales::alpha("white", 0.85),
           colour = "#2C5F8A",
           linewidth = 0.25) +
  scale_y_continuous(limits = c(-0.25, 0.25)) +
  theme_minimal(base_size = 7.2) +
  labs(
    title = "Macroeconomic Drift & Vintage Trends",
    subtitle = "Annual association: 14.0% lower conditional default odds per later origination year",
    x = "Loan Origination Date (2007 - 2011)",
    y = "Monthly Baseline Deviation"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 7.8, color = "midnightblue"),
    plot.subtitle = element_text(size = 6.2, color = "gray30"),
    axis.title.x = element_text(size = 6.6, face = "bold"),
    axis.title.y = element_text(size = 6.6, face = "bold"),
    plot.margin = margin(t = 2, r = 2, b = 2, l = 2, unit = "pt")
  )

p2_state + p2_time + plot_layout(widths = c(1, 1.25))
```

* **The "Ranking vs. Monitoring" Finding:** On the 2011 temporal holdout, adding the fitted state and time effects did not materially improve applicant rank discrimination: GLM Gini **0.4070** versus GLMM Gini **0.4069**. Borrower-level characteristics therefore provided essentially all of the rank-ordering performance achieved by the evaluated models on this holdout.
* **Recommended Operational Role:** Advance the parsimonious GLM as the candidate applicant-ranking architecture for local redevelopment, recalibration, and independent validation. Use the GLMM for historical vintage monitoring, geographic portfolio decomposition, and identifying cohorts requiring investigation.
* **Fair Lending & Governance Guardrails:** Nevada, Florida, and California had positive estimated baseline deviations with approximate conditional intervals excluding zero (state variance represents $0.88\%$ of latent variance). These exploratory historical associations do not establish causality. **Management must not implement regional rejection quotas or geographic risk pricing**, which could create material fairness, legal, and governance concerns.
* **Temporal Trend Association:** Holding observed borrower characteristics constant, later origination years were associated with lower default odds ($\text{OR} = 0.860$, 95% CI: $[0.798, 0.927]$). Possible contributors include underwriting changes, borrower composition, outcome seasoning, and broader historical conditions.

# Diagnostic Findings, Governance Limitations & Deployment Roadmap
To support model-governance review consistent with credit-risk management expectations, we evaluated model diagnostics and operational requirements:

* **Diagnostic Findings:**
  - *Global Checks:* Overall residual uniformity ($p = 0.811$) and dispersion ($p = 0.756$) tests were non-significant; adjusted GVIF values below 1.6 indicate no severe multicollinearity likely to destabilize fitted main-effect estimates.
  - *Qualifications:* Grouped risk-band dispersion ($p = 0.020$) identified uneven variability across coarse predicted-risk groups, with mild underprediction in several intermediate-to-high risk bands; monthly aggregate residuals retained positive serial dependence ($DW = 0.984$, $p < 0.001$), indicating persistence between adjacent cohorts.
* **Prudential Roadmap for Australian Banking Deployment:**
  - *Representative Australian Data:* Before deployment, redevelop or recalibrate the model using representative Australian application, repayment, credit-bureau, affordability, and macroeconomic data, followed by independent validation and fairness assessment.
  - *Complete Expected Credit Loss Framework:* The model provides Probability of Default (PD) only. Capital and expected-loss applications require validated PD, Loss Given Default (LGD), and Exposure at Default (EAD) components ($\text{ECL} = \text{PD} \times \text{LGD} \times \text{EAD}$) under approved governance.
